Paradigms in Physics 2.0

### 2017-06-28 paper.tex
Update on Overleaf.

%\documentclass[english,aps,prstper,reprint,showpacs]{revtex4-1}   % This version should work (prstper option), but actually formats reference numbers incorrectly.
\documentclass[english,aps,pra,reprint,noshowpacs,superscriptaddress]{revtex4-1}   
% Using pra instead of prstper gives correct square-bracket (not superscript) reference formatting
\usepackage[T1]{fontenc}	% should generally be included for better accented-word behavior
\usepackage[latin9]{inputenc}	% should generally be included for better accent behavior
\usepackage{geometry}		% for controlling page margins
\geometry{verbose,tmargin=1in,bmargin=1in,lmargin=0.75in,rmargin=0.75in}	% define margins
\usepackage{graphicx}
\usepackage[above,below]{placeins}	% allows use of \FloatBarÂ­rier command to force section barriers
\usepackage{times}
% Next three lines are optional, use the hyperref package to make URLs and reference links live.
\usepackage{hyperref}  
\hypersetup{colorlinks=true,urlcolor=blue,citecolor=blue,linkcolor=blue}   
\urlstyle{same}
\pagestyle{empty}			% page numbers added later, when compiling the whole proceedings
\begin{document}

\title{Paradigms in Physics 2.0}
\author{David Roundy}
\author{Liz Gire}
\author{Ethan Minot}
\author{Emily van Zee}
\affiliation{Department of Physics, Oregon State University, Corvallis, Oregon, 97331}
\author{Tevian Dray}
\affiliation{Department of Mathematics, Oregon State University, Corvallis, Oregon, 97331}
\author{Corinne A. Manogue}
\affiliation{Department of Physics, Oregon State University, Corvallis, Oregon, 97331}

@@

\begin{abstract}
In 2016, the Department of Physics at Oregon State University began a
process to revise our Paradigms in Physics curriculum for physics
majors.  This poster will describe both the process by which our
department achieved consensus on this curricular change, and the
resulting curriculum.  The Paradigms 2.0 committee begain with a
survey of students and faculty, followed by individual interviews with
the faculty teaching each existing course.  As we developed a plan to
address student- and faculty-identified challenges in the curriculum,
we met with each faculty member individually to explain and refine our
proposal, which was unanimously approved by the faculty.  Major
changes include elimination or major changes to several courses (math
methods, computational physics, modern physics, electronics, and
classical mechanics), including the introduction of two sophomore-year
courses designed specifically to help prepare students to for their
upper-division courses.
\end{abstract}

\maketitle

\section{Introduction}
The Paradigms in Physics program is a rethinking of the upper-division
physics curriculum that began around the time when our current
students were born~\cite{manogue2001paradigms}.  The basic ideas
behind the original Paradigms reform were (1) attention to the
sequencing of concepts, (2) acheiving faculty buy-in, which requires
faculty to understand the sequencing, and (3) implementation of active
engagement through the upper-division curriculum.  In 2016, we began a
rethinking of the Paradigms, with focus on the same principles guiding
the original effort, but with the input of new faculty, and applying
lessons learned over the past two decades.  In this paper, we will
explain the reasoning behind this Paradigms 2.0 effort, the process by
which we achieved developed and reached consensus on an improved
curriculum, and will give both a summary and a few highlights of the
new program.

\subsection{Background}
The Paradigms in Physics project began in 1996, when three faculty
members at Oregon State applied to the NSF for funding to redesign the
upper-division curriculum.  The motivation for the original Paradigms
project was in fact similar to the motivation for Paradigms 2.0: a
desire to soften the ``brick wall'' encountered by students upon
reaching the junior year, within the constraint that transfer students
must be able to graduate with two years of upper-division
courses~\cite{manogue2001paradigms}.  On top of this, there was a
desire to provide students with a broad knowledge of physics prior to
the GRE exams in the fall of their senior year.  The process of
redesigning the curriculum was dominated by the creation index cards
listing subject content, and sorting these cards into courses.  This
process involved the entire faculty, and culminated in a unanimous
agreement to adopt the resulting curriculum.

@@

A coupe of decades have passed since the original Paradigms effort.
During this time we have made a number of changes: e.g. a new
\emph{first} Paradigm was introduced to soften the beginning of the
junior year, the order of Paradigms was shuffled more than once, Math
Methods was moved from the beginning of the senior year to the end of
the junior year, and a computational laboratory course was introduced
to accompany the Paradigms.  We maintained the tradition of meeting
every three weeks to discuss issues relating to upper-division
teaching.  New faculty arrived and learned to teach the new courses,
and introduced their own approach to the courses.

\subsection{Motivation for Paradigms 2.0}
There were a number of factors that motivated us to embark on the
Paradigms~2.0 process.  In the last two decades, we have observed
challenges students face in the Paradigms.  While in many cases we
addressed these by changing and reordering the courses, we felt a look
at the entire curriculum was in order.  In addition, our faculty
buy-in and aware was diminishing by attrition: while our younger
faculty are enthusiastic about the Paradigms, many lack perspective on
where students are in a given course, and what content is essential
for students to grasp prior to a following course.

\paragraph{Transfer students}
Our curriculum is strongly affected by our desire to accomodate
transfer students from community colleges.  This leads us to focus on
a physics major in which all upper-division courses are taken in two
years.  However, our current curriculum is very hard on students in
the fall of their junior year, and especially so for transfer
students.  So we aimed to structure the courses such that non-transfer
students could be better prepared for their junior year, while
transfer students could postpone a few courses so that their junior
Fall would be no more difficult than that of any other student.

\paragraph{Coupled curricular changes}
A backlog of curricular changes have accumulated that could not be
separated from an examination of the curriculum as a whole.  Most
notable were discussions of the requirements for computational
physics, the amount of required electronics credits, and the role of
our modern physics course.  These persistent issues were not easily
addressed without addressing the physics major curriculum as a whole,
and any discussion of one tended to devolve into a discussion of
everything, without progress being made.

@@

The committee met twice a week to discuss existing challenges and sort
the cards into new stacks corresponding to new and reordered courses.
When we discussed major changes to a course, we often invited
interested faculty to join us to provide their own perspective on
possible challenges and improvements.  Once we had a draft proposal,
we began inviting each faculty member in to see the cards and discuss
the proposed sequence of courses.  In addition, we had a meeting with
all the current students to explain the proposed changes and request
feedback.

When we had obtained feedback from the entire faculty, we scheduled
\emph{two} full faculty meetings.  In the first meeting, we presented
our proposal in detail and invited questions, but not discussion.
This was needed for several reasons.  Those faculty we met with first
may not have seen the final proposal.  Some faculty during their
individual meetings chose to focus on a small subset of the
curriculum, often relating to courses they had themselves taught.  In
one meeting we presented the proposal and fielded questions.  In a
second faculty meeting, we again presented our proposal, and opened
the floor for discussion.  After considerable discussion, we
unanimously voted to adopt the proposed changes.

@@

\subsection{5-week Paradigms}
To begin with the most distinctive feature of our curriculum, we
wanted to maintain the existing intensive 7-hour-per-week schedule in
the junior year, which we have found effective.  However, we chose to
change from three 3-week Paradigms per quarter to two 5-week Paradigms
per quarter.    This change has several advantages.  It gives students
a bit more time with each professor prior to their final exam.  This
gives a bit more flexibility, for instance, in case of illness.
Finally, by reducing the number of Paradigms, we allow the curriculum
to be taught with fewer faculty, albeit at a work load per professor.

We put considerable thought into the content and ordering of the new
Paradigms.  In most cases, we think of each 5-week Paradigm as either
one formerly 4-week Paradigm (the one that consumed the extra week
each quarter) or two 3-week Paradigms compressed.  The major
scheduling change is to place Static Fields (electrostatics and
magnetostatics) in the spring quarter.  This intended to give students
a bit more time to take Vector Calculus.  It also puts all use of
curvilinear coordinates in the spring quarter, which should ease the
learning of Central Forces.  The final major change was the
elimination of special relativity from the Paradigms, which is has now
moved into the new sophomore-level Theoretical Mechanics course.

\subsection{Math bits}
We chose to eliminate the Math Methods course in favor of formal
``math bits'' in the Paradigms.  Our previous Math Methods Capstone
posed a challenge regarding placement in the curriculum: it needed to
be sufficiently late in order for students to be adequately prepared
for advanced mathematical content, but in order to be functional in
the curriculum it should teach content that is required by other
physics courses, and thus should be earlier in the curriculum.  In
response to these issues, and in recognition that the majority of our
majors are not bound for graduate school, and have no need for
advanced math methods, we chose to integrate the needed math content
into the Paradigms, and allow advanced students to take our
graduate-level Math Methods course.  In order to ensure that our
students to receive the math instruction they require, and based on
our previous experience with math-intensive weeks (the ``odd'' weeks)
in the Paradigms, we chose to integrate one week of ``math bits'' into
each Paradigm.  The math bits for the entire year are treated as a
distinct teaching assignment, so students will be accustomed to the
treatment of math as a distinct subject supporting the physics
content.  This provides some relief from teaching for the Paradigms
professors, and at the same time ensures that they do not succumb to
the temptation to short-change the math content to the detriment of
the students.

@@

A final change to Electronics is that we now will \emph{require}
Electronics during the junior year (specifically as a prerequisite for
Oscillations and Waves).  This is made possible by the reduction in
Fall workload for incoming transfer students, who had previously
usually taken Electronics as a senior.  This change has enabled us to
articulate distinct and sequenced learning outcomes from these two
courses, particularly in the realm of complex exponentials and Fourier
transforms.  It is also beneficial in providing students with
trouble-shooting skills prior to the in-class labs taught in
Oscillations and Waves.

\subsection{Computational lab}
Over the last six years, we have been developing a 1-credit
computational laboratory course that accompanies the Paradigms.  We
chose to require this course, while removing a requirement for a
lower-division 3-credit course in computational physics.  This
lower-division course was challenging to teach, since it always had a
mix of lower- and upper-division students, with very different skill
levels and needs.  Transfer students now take computation alongside
the non-transfer students, in a laboratory setting using pair
programming to help new programmers to learn.

\subsection{Sophomore courses}
We introduced two new sophomore-level courses: \emph{Challenges in
  Contemporary Physics}, and \emph{Theoretical Mechanics}.  Both of
these courses ramp up student mathematical abilities prior to their
junior year.  The Challenges course will focus on estimation,
dimensional reasoning, and interpretation of integrals, while
Theoretical Mechanics will teach power series approximations, and
expose students to increased levels of mathematical sophistication.

@@

\subsubsection{Challenges in Contemporary Physics}

>>>>

%\documentclass[english,aps,prstper,reprint,showpacs]{revtex4-1}   % This version should work (prstper option), but actually formats reference numbers incorrectly.
\documentclass[english,aps,pra,reprint,noshowpacs,superscriptaddress]{revtex4-1}   
% Using pra instead of prstper gives correct square-bracket (not superscript) reference formatting
\usepackage[T1]{fontenc}	% should generally be included for better accented-word behavior
\usepackage[latin9]{inputenc}	% should generally be included for better accent behavior
\usepackage{geometry}		% for controlling page margins
\geometry{verbose,tmargin=1in,bmargin=1in,lmargin=0.75in,rmargin=0.75in}	% define margins
\usepackage{graphicx}
\usepackage[above,below]{placeins}	% allows use of \FloatBarÂ­rier command to force section barriers
\usepackage{times}
% Next three lines are optional, use the hyperref package to make URLs and reference links live.
\usepackage{hyperref}  
\hypersetup{colorlinks=true,urlcolor=blue,citecolor=blue,linkcolor=blue}   
\urlstyle{same}
\pagestyle{empty}		% page numbers added later, when compiling the whole proceedings

\newcommand\corinnesays[1]{\textcolor{red}{[\sc CAM: \em{#1}]}}

\begin{document}

\title{Paradigms in Physics 2.0}
\author{David Roundy}
\author{Liz Gire}
\author{Ethan Minot}
\author{Emily van Zee}
\affiliation{Department of Physics, Oregon State University, Corvallis, Oregon, 97331}
\author{Corinne A. Manogue}
\affiliation{Department of Physics, Oregon State University, Corvallis, Oregon, 97331}

@@

\begin{abstract}
In 2016, the Department of Physics at Oregon State University began a
process to revise our Paradigms in Physics curriculum for physics
majors.  This paper will describe both the process by which our
department achieved consensus on this curricular change, and the
resulting curriculum.  The Paradigms 2.0 committee begain with a
survey of students and faculty, followed by individual interviews with
the faculty teaching each existing course.  As we developed a plan to
address student- and faculty-identified challenges in the curriculum,
we met with each faculty member individually to explain and refine our
proposal, which was unanimously approved by the faculty.  Major
changes include elimination or major changes to several courses (math
methods, computational physics, modern physics, electronics, and
classical mechanics), including the introduction of two sophomore-year
courses designed specifically to help prepare students to for their
upper-division courses.
\end{abstract}

\maketitle
% * <ethan.minot@oregonstate.edu> 2017-06-28T19:21:10.109Z:
%
% ^.
\section{Introduction}
The Paradigms in Physics program is a rethinking of the upper-division
The Paradigms in Physics project began in 1996, when three faculty
members at Oregon State applied to the NSF for funding to redesign the
upper-division curriculum.  The motivation for the original Paradigms
project was in fact similar to the motivation for Paradigms 2.0: a
desire to soften the ``brick wall'' encountered by students upon
reaching the junior year, within the constraint that transfer students
must be able to graduate with two years of upper-division
courses~\cite{manogue2001paradigms}.  On top of this, there was a
desire to provide students with a broad knowledge of physics prior to
the GRE exams in the fall of their senior year.  The process of
redesigning the curriculum was dominated by the creation of index cards
listing subject content, and sorting these cards into courses.  This
process involved the entire faculty, and culminated in a unanimous
agreement to adopt the resulting curriculum.

@@

Two decades have passed since the original Paradigms effort.
During this time we have made a number of changes: e.g. a new
\emph{first} Paradigm was introduced to soften the beginning of the
junior year, the order of Paradigms was shuffled more than once, Math
Methods was moved from the beginning of the senior year to the end of
the junior year, and a computational laboratory course was introduced
to accompany the Paradigms.  The faculty maintained the tradition of meeting
every three weeks to discuss issues relating to upper-division
teaching.  New faculty arrived and learned to teach the new courses,
and introduced their own ideas to the courses.

\subsection{Motivation for Paradigms 2.0}
There were a number of factors that motivated us to embark on the
Paradigms~2.0 process.  In the last two decades, we have observed
a number of challenges students face in the Paradigms.  While in many cases we
addressed these by changing and reordering the courses, we felt a look
at the entire curriculum was in order.  In addition, our faculty
understanding of the details of the sequence was diminishing by attrition: 
while our younger
faculty are enthusiastic about the Paradigms, many lack perspective on
where students are in a given course, and what content is essential
for students to grasp prior to a following course.

\paragraph{Transfer students}
Our curriculum is strongly affected by our desire to accomodate
transfer students from community colleges.  This leads us to focus on
a physics major in which all upper-division courses are taken in two
years.  However, our current curriculum is very hard on students in
the fall of their junior year, and especially so for transfer
students.  So we aimed to structure and order the courses such that non-transfer
students could be better prepared for their junior year, while
transfer students could postpone a few courses so that their junior
Fall would be no more difficult than that of any other student.

\paragraph{Coupled curricular changes}
A backlog of curricular changes have accumulated that could not be
separated from an examination of the curriculum as a whole.  Most
notable were discussions of the requirements for computational
physics, the amount of required electronics credits, and the role of
our modern physics course.  These persistent issues were not easily
addressed without addressing the physics major curriculum as a whole,
and any discussion of one tended to devolve into a discussion of
everything, without progress being made.  
\corinnesays{This paragraph could perhaps be tightened.}

@@

The committee met twice a week to discuss existing challenges and sort
the cards into new stacks corresponding to new and reordered courses.
When we discussed major changes to a course, we often invited
interested faculty to join us to provide their own perspective on
possible challenges and improvements.  Once we had a draft proposal,
we began inviting each faculty member in to see the cards and discuss
the proposed sequence of courses.  In addition, we had a focus group with
all the current students to explain the proposed changes and request
feedback.

When we had obtained feedback from the entire faculty, we scheduled
\emph{two} full faculty meetings.  In the first meeting, we presented
our proposal in detail and invited questions, but not discussion.
This was needed for several reasons.  Those faculty we met with first
may not have seen the final proposal.  Some faculty during their
individual meetings chose to focus on a small subset of the
curriculum, often relating to courses they had themselves taught.  In a
second faculty meeting, we again presented our proposal, and opened
the floor for discussion.  After considerable discussion, we
unanimously voted to adopt the proposed changes.

@@

\subsection{5-week Paradigms}
To begin with the most distinctive feature of our curriculum, we
wanted to maintain the existing intensive 7-hour-per-week schedule in
the junior year, which we have found effective.  However, we chose to
change from three 3-week Paradigms per quarter to two 5-week Paradigms
per quarter.    This change has several advantages.  It gives students
a bit more time with each professor prior to their final exam.  This
gives a bit more flexibility, for instance, in case of illness.
Finally, by reducing the number of Paradigms, we allow the curriculum
to be taught with fewer faculty, albeit at an increased work load per professor.

We put considerable thought into the content and ordering of the new
Paradigms.  In most cases, we think of each 5-week Paradigm as either
one formerly 4-week Paradigm (the one that consumed the extra week
each quarter), two 3-week Paradigms compressed, or three 3-week paradigms combined into two 5-week paradigms.  The major
scheduling change is to place Static Fields (electrostatics and
magnetostatics) in the spring quarter.  
This intended to give students, especially transfer students, 
a bit more time to take Vector Calculus.  It also puts all use of
curvilinear coordinates in the spring quarter, which should ease the
learning of Central Forces.  The final major change was to move special relativity from the junior-year Paradigms to the new sophomore-level Theoretical Mechanics course.

\subsection{Math bits}
We chose to eliminate the Math Methods course in favor of formal
``math bits'' in the Paradigms.  Our previous Math Methods Capstone
posed a challenge regarding placement in the curriculum: it needed to
be sufficiently late for students to be adequately prepared
for advanced mathematical content, but in order to be functional in
the curriculum it should teach content that is required by other
physics courses, and thus should be earlier in the curriculum.  In
response to these issues, and in recognition that the majority of our
majors are not bound for graduate school, and have no need for
advanced math methods, we chose to integrate the needed math content
into the Paradigms, and allow advanced students to take our
graduate-level Math Methods course.  In order to ensure that our
students to receive the math instruction they require, and based on
our previous experience with math-intensive weeks (the ``odd'' weeks)
in the Paradigms, we chose to integrate one week of ``math bits'' into
each Paradigm.  The math bits for the entire year are treated as a
distinct teaching assignment, so students will be accustomed to the
treatment of math as a distinct subject supporting the physics
content.  This provides some relief from teaching for the Paradigms
professors, and at the same time ensures that they do not succumb to
the temptation to short-change the math content to the detriment of
the students.
\corinnesays{This paragraph is awkward.  I'm still thinking about some rewording.  This might be a good paragraph to discuss this afternoon.}

@@

A final change to Electronics is that we now will \emph{require}
Electronics during the junior year (specifically as a prerequisite for
the Oscillations and Waves Paradigm).  This is made possible by the reduction in
Fall workload for incoming transfer students, who had previously
usually taken Electronics as a senior.  This change has enabled us to
articulate distinct and sequenced learning outcomes from these two
courses, particularly in the realm of complex exponentials and Fourier
transforms.  It is also beneficial in providing students with
trouble-shooting skills prior to the in-class labs taught in
Oscillations and Waves course.

\subsection{Computational lab}
Over the last six years, we have been developing a 1-credit
computational laboratory course that accompanies the Paradigms.  We
chose to require this course, while removing a requirement for a
lower-division 3-credit course in computational physics.  This
lower-division course was challenging to teach, since it always had a
mix of lower- and upper-division students, with very different skill
levels and needs.  Transfer students now take computation alongside
the non-transfer students. The course is taught in a laboratory setting using pair
programming to help new programmers to learn.

\subsection{Sophomore courses}
We introduced two new sophomore-level courses: \emph{Challenges in
  Contemporary Physics}, and \emph{Theoretical Mechanics}.  Both of
these courses ramp up student mathematical abilities prior to their
junior year.  The Challenges course focuses on estimation,
dimensional reasoning, and interpretation of integrals, while
Theoretical Mechanics will teach power series approximations, and
expose students to increased levels of mathematical sophistication.

@@

\subsubsection{Challenges in Contemporary Physics}
\corinnesays{You might want to use your paragraph structure here, rather than sub-sub sections.}
